---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fda)
library(tidyverse)
library(plotly)
library(dplyr)
library(fdapace)
# library(aplpack)
library(ks)
library(devtools)
library(ggfortify)
```

Data wrangle, preparation and visualization:

```{r}

dt.1 <-  sort(sample(nrow(AP_GRF_stance_N), nrow(AP_GRF_stance_N)*.7))
train.AP<-AP_GRF_stance_N[dt.1,]
test.AP<-AP_GRF_stance_N[-dt.1,]

train.ML<-ML_GRF_stance_N[dt.1,]
test.ML<-ML_GRF_stance_N[-dt.1,]

train.V<-V_GRF_stance_N[dt.1,]
test.V<-V_GRF_stance_N[-dt.1,]

train.COPx<-COPx_stance[dt.1,]
test.COPx<-COPx_stance[-dt.1,]

train.COPy<-COPy_stance[dt.1,]
test.COPy<-COPy_stance[-dt.1,]

train.ID <-ID_info[dt.1,]
test.ID <-ID_info[-dt.1,]


data_pre_func <- function(df1, df2){
  df <- cbind(df1, df2)
  df3 <- df %>% pivot_longer(-c(ID, KNEE, TRIAL, tr_length), names_to = "time", values_to = "force")%>%
    mutate(new = paste(ID, "-",KNEE, "-",TRIAL)) 
  
  t <- rep(seq(0,100,length.out = 100),nrow(df))
  
  df_pre <- cbind(df3, t)
  return(df_pre)
}

try <- data_pre_func(train.ID, train.AP)
try2 <- data_pre_func(train.ID, train.ML)
try3 <- data_pre_func(train.ID, train.V)
try4 <- data_pre_func(train.ID, train.COPx)
try5 <- data_pre_func(train.ID, train.COPy)

# p_try <- ggplot(try, aes(x = t, y = force, group = new, col = new)) +
#   geom_line()
# p_try
```

Do FPCA for all 5 time normalized dfs:
```{r}
L_AP <- MakeFPCAInputs(try$new, try$t, try$force)
FPCAdense_AP <- FPCA(L_AP$Ly, L_AP$Lt)

# Plot the FPCA object
plot(FPCAdense_AP)


# Select the first PCs:
AP_stance_dr <- FPCAdense_AP$xiEst
colnames(AP_stance_dr) <- c('AP_FPC1', 
                            'AP_FPC2', 
                            'AP_FPC3',
                            'AP_FPC4',
                            'AP_FPC5', 
                            'AP_FPC6', 
                            'AP_FPC7',
                            'AP_FPC8',
                            'AP_FPC9',
                            'AP_FPC10',
                            'AP_FPC11',
                            'AP_FPC12')

```


```{r}
L_ML <- MakeFPCAInputs(try2$new, try2$t, try2$force)
FPCAdense_ML <- FPCA(L_ML$Ly, L_ML$Lt)

# Plot the FPCA object
plot(FPCAdense_ML)


# Select the first PCs:
ML_stance_dr <- FPCAdense_ML$xiEst
colnames(ML_stance_dr) <- c('ML_FPC1', 
                            'ML_FPC2', 
                            'ML_FPC3',
                            'ML_FPC4',
                            'ML_FPC5', 
                            'ML_FPC6', 
                            'ML_FPC7',
                            'ML_FPC8',
                            'ML_FPC9',
                            'ML_FPC10',
                            'ML_FPC11',
                            'ML_FPC12',
                            'ML_FPC13',
                            'ML_FPC14',
                            'ML_FPC15')

```


```{r}
L_V <- MakeFPCAInputs(try3$new, try3$t, try3$force)
FPCAdense_V <- FPCA(L_V$Ly, L_V$Lt)

# Plot the FPCA object
plot(FPCAdense_V)


# Select the first PCs:
V_stance_dr <- FPCAdense_V$xiEst
colnames(V_stance_dr) <- c('V_FPC1', 
                            'V_FPC2', 
                            'V_FPC3',
                            'V_FPC4',
                            'V_FPC5', 
                            'V_FPC6', 
                            'V_FPC7',
                            'V_FPC8')

```


```{r}
L_COPx <- MakeFPCAInputs(try4$new, try4$t, try4$force)
FPCAdense_COPx <- FPCA(L_COPx$Ly, L_COPx$Lt)

# Plot the FPCA object
plot(FPCAdense_COPx)


# Select the first PCs:
COPx_dr <- FPCAdense_COPx$xiEst
colnames(COPx_dr) <- c('COPx_FPC1', 
                       'COPx_FPC2', 
                       'COPx_FPC3',
                       'COPx_FPC4',
                       'COPx_FPC5')

```


```{r}
L_COPy <- MakeFPCAInputs(try5$new, try5$t, try5$force)
FPCAdense_COPy <- FPCA(L_COPy$Ly, L_COPy$Lt)

# Plot the FPCA object
plot(FPCAdense_COPy)


# Select the first PCs:
COPy_dr <- FPCAdense_COPy$xiEst
colnames(COPy_dr) <- c('COPy_FPC1', 
                       'COPy_FPC2')


GRF_FPCs <- cbind(AP_stance_dr, ML_stance_dr, V_stance_dr)
COP_FPCs <- cbind(COPx_dr, COPy_dr)
FPCs <- cbind(GRF_FPCs, COP_FPCs)
colnames(FPCs)
```



Based on the scree-plot we see that the first three components appear to encapsulate most of the relevant variation. The number of eigen components to reach a 99% FVE is 12 but just 7 eigen components are enough to reach a 95.0%.

AP: 3 5 7 12
ML: 3 5 7 15
V: 2 2 4 8
COPx: 1 1 2 5
COPy: 1 1 1 2

```{r}
SelectK(FPCAdense_AP, criterion = "FVE", FVEthreshold = 0.8)
SelectK(FPCAdense_AP, criterion = "FVE", FVEthreshold = 0.9)
SelectK(FPCAdense_AP, criterion = 'FVE', FVEthreshold = 0.95)
SelectK(FPCAdense_AP, criterion = 'FVE', FVEthreshold = 0.99)

SelectK(FPCAdense_ML, criterion = "FVE", FVEthreshold = 0.8)
SelectK(FPCAdense_ML, criterion = "FVE", FVEthreshold = 0.9)
SelectK(FPCAdense_ML, criterion = 'FVE', FVEthreshold = 0.95)
SelectK(FPCAdense_ML, criterion = 'FVE', FVEthreshold = 0.99)

SelectK(FPCAdense_V, criterion = "FVE", FVEthreshold = 0.8)
SelectK(FPCAdense_V, criterion = "FVE", FVEthreshold = 0.9)
SelectK(FPCAdense_V, criterion = 'FVE', FVEthreshold = 0.95)
SelectK(FPCAdense_V, criterion = 'FVE', FVEthreshold = 0.99)

SelectK(FPCAdense_COPx, criterion = "FVE", FVEthreshold = 0.8)
SelectK(FPCAdense_COPx, criterion = "FVE", FVEthreshold = 0.9)
SelectK(FPCAdense_COPx, criterion = 'FVE', FVEthreshold = 0.95)
SelectK(FPCAdense_COPx, criterion = 'FVE', FVEthreshold = 0.99)

SelectK(FPCAdense_COPy, criterion = "FVE", FVEthreshold = 0.8)
SelectK(FPCAdense_COPy, criterion = "FVE", FVEthreshold = 0.9)
SelectK(FPCAdense_COPy, criterion = 'FVE', FVEthreshold = 0.95)
SelectK(FPCAdense_COPy, criterion = 'FVE', FVEthreshold = 0.99)
```


```{r}
#standardize FPCs matrix:
mat.Out <-  scale(FPCs)
train.ID$ID <- as.character(train.ID$ID)
train.ID$TRIAL <- as.character(train.ID$TRIAL)
ddd <- cbind(train.ID, mat.Out)
ddd <- ddd %>% mutate(new = paste(ID, "-",KNEE, "-",TRIAL))

aaa <- cbind(train.ID, FPCs)
aaa <- aaa %>% mutate(new = paste(ID, "-",KNEE, "-",TRIAL))

train.dis <-discrete[dt.1,]
test.dis <-discrete[-dt.1,]


FPCs.pca <- prcomp(FPCs[, c(1:3, 13:15, 28:29, 36,41)], center = TRUE,scale. = TRUE)
summary(FPCs.pca)

FPCs.all.pca <- prcomp(FPCs, center = TRUE,scale. = TRUE)
summary(FPCs.all.pca)

library(devtools)
library(ggfortify)

autoplot(FPCs.pca,
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3,
         data = aaa,
         colour = 'TRIAL')


ss <- kmeans(FPCs, 4)

set.seed(1)
autoplot(kmeans(FPCs, 3),frame = TRUE, data = FPCs)

autoplot(kmeans(train.dis, 3),frame = TRUE, data = train.dis)

set.seed(2)
autoplot(kmeans(FPCs[, c(1:3, 13:15, 28:29, 36,41)], 3),frame = TRUE, data = FPCs)

set.seed(3)
autoplot(kmeans(mat.Out, 4),frame = TRUE, data = mat.Out)

set.seed(4)
autoplot(kmeans(mat.Out[, c(1:3, 13:15, 28:29, 36,41)], 4), data = mat.Out)

```



```{r}
par(mfrow=c(1,2))
  CreatePathPlot(FPCAdense_AP,subset = c(3,5,135), K = 12, main = 'K = 12', pch = 4); grid()
  CreatePathPlot(FPCAdense_AP, subset = c(3,5,135), K = 7, main = 'K = 7', pch = 4) ; grid()
  
  
par(mfrow=c(1,2))
  CreatePathPlot(FPCAdense_ML,subset = c(3,5,135), K = 15, main = 'K = 15', pch = 4); grid()
  CreatePathPlot(FPCAdense_ML, subset = c(3,5,135), K = 7, main = 'K = 7', pch = 4) ; grid()
```

perform outlier detection:

```{r}
par(mfrow=c(1,1))
CreateOutliersPlot(FPCAdense_AP, optns = list(K = 12, variant = 'KDE'))

par(mfrow=c(1,1))
  CreateOutliersPlot(FPCAdense_ML, optns = list(K = 15, variant = 'KDE'))
  
par(mfrow=c(1,1))
  CreateOutliersPlot(FPCAdense_V, optns = list(K = 8, variant = 'KDE'))
  
par(mfrow=c(1,1))
  CreateOutliersPlot(FPCAdense_COPx, optns = list(K = 5, variant = 'KDE'))
  
par(mfrow=c(1,1))
  CreateOutliersPlot(FPCAdense_COPy, optns = list(K = 2, variant = 'KDE'))
  
```


visualize data using a functional box-plot
```{r}
# CreateFuncBoxPlot(FPCAdense, xlab = 'Stance time', ylab = 'VGRF force', optns = list(K = 7, variant='KDE'))
```

```{r}
# library(EMCluster)
# FClust(L_try$Ly, L_try$Lt, k = 4, cmethod = "EMCluster", optnsFPCA = NULL, optnsCS = NULL)
```


```{r}
# library(corrplot)
# corrplot(FPCAdense$fittedCorr, method="color")
```





